## Process ONT UMI data from Zymomock rRNA amplicons

**Reference**

* SM Karst, RM Ziels, RH Kirkegaard, EA Sørensen, D. McDonald, Q Zhu, R Knight, & M Albertsen. (2020). Enabling high-accuracy long-read amplicon sequences using unique molecular identifiers with Nanopore or PacBio sequencing. [bioRxiv, 6459039](https://www.biorxiv.org/content/10.1101/645903v3).

**Sample**

* ZymoBIOMICS Microbial Community DNA Standard (D6306, lot no. ZRC190811, Zymo Research)
* 8 bacterial species: Bacillus subtilis, Enterococcus faecalis, Escherichia coli, Lactobacillus fermentum, Listeria monocytogenes, Pseudomonas aeruginosa, Salmonella enterica, Staphylococcus aureus.

**Library**

* [UMI amplicon protocol at Protocols.io](https://www.protocols.io/private/F5C5FE21305911EAAC0B0242AC110003)
* rRNA PCR primers: [8F] 5'-AGRGTTYGATYMTGGCTCAG and [1391R] 5'-GACGGGCGGTGWGTRCA
* Library: 1D amplicon/cDNA by ligation (SQK-LSK109)

**Sequencing**

* Flowcell: R10
* Instrument: MinION
* Basecalling: guppy v3.4.4

### Generate UMI consensus sequences

Download and decompress data
```{bash eval=FALSE}
wget "ftp://ftp.sra.ebi.ac.uk/vol1/run/ERR381/ERR3813594/smkj412_zymo_D6306_rrna_umi_ont_min_r10_g344hac.fq.gz"
gunzip -c smkj412_zymo_D6306_rrna_umi_ont_min_r10_g344hac.fq.gz > reads.fq
```

Generate UMI consensus sequences from raw read data
```{bash eval=FALSE}
longread_umi nanopore_pipeline \
  -d reads.fq \
  -o umi_out \
  -v 25 \
  -q r10_min_high_g340 \
  -m 3500 \
  -M 6000 \
  -s 90 \
  -e 90 \
  -f CAAGCAGAAGACGGCATACGAGAT \
  -F AGRGTTYGATYMTGGCTCAG \
  -r AATGATACGGCGACCACCGAGATC \
  -R CGACATCGAGGTGCCAAAC \
  -c 2 \
  -p 2 \
  -t 40 \
  -T 1
```

Generate validation data from UMI consensus sequences
```{bash eval=FALSE}
longread_umi qc_pipeline \
-d "reads.fq" \
-c "umi_out/consensus_raconx2_medakax2.fa;umi_out/variants.fa" \
-r "zymo_curated" \
-u umi_out \
-o umi_out/qc \
-t 40
```

## Validate UMI consensus sequences

Load longread_umi functions and create working directory
```{r eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
source("longread_umi/scripts/validation_functions.R")
# Conda path: $CONDA_PREFIX/longread_umi/scripts/validation_functions.R"
dir.create("./validation")
```

Compile validation data and generate error profile
```{r eval=FALSE, echo=TRUE, message=FALSE, warning=FALSE}
lu_compile_qc(
  data_dir = "umi_out/qc",
  umi_consensus = "consensus_raconx2_medakax2.fa",
  reference = "zymo-ref-uniq_2019-10-28.fa",
  read_orientation = "read_orientation.txt", 
  silva = NULL,
  out_path = "validation/ONT_R10_ZYMO_rRNA_qc.Rdata"
)
load("validation/ONT_R10_ZYMO_rRNA_qc.Rdata")

ep <- lu_error_profile(
  refs = "umi_out/qc/zymo-ref-uniq_2019-10-28.fa",
  sam = "umi_out/qc/consensus_raconx2_medakax2.sam"
)
save(ep, file = "validation/ONT_R10_ZYMO_rRNA_ep.Rdata")
```


Or load pregenerated data (e.g. longread_umi/docs/rdata/)
```{r echo=TRUE, message=FALSE, warning=FALSE}
load("validation/ONT_R10_ZYMO_rRNA_qc.Rdata")
load("validation/ONT_R10_ZYMO_rRNA_ep.Rdata")
```


Perform UMI bin post-processing filtering (will be incoporated in pipeline in the future)
```{r echo=TRUE, message=FALSE, warning=FALSE}
qc <- filter(
  qc,
  ror > 10^-0.6, # remove bins with extreme read orientation ratios
  ror < 10^0.6,
  umi_bin_size/umi_cluster_size < 10, # remove bins with extreme bin to cluster size
  umi_match_error < 3.5, # remove bins with high mismatch between UMI reference and reads
  !is.na(length) # remove umi bins where >=1 gene specific primers was not found in consensus sequence.
)

ep <- filter(
  ep,
  sub(";.*","", qname) %in% qc$umi
)
```


### Basic data characteristics

Plot read size distribution.
```{r echo=TRUE, message=FALSE, warning=FALSE}
lu_plot_hist(
  value = read_lengths, 
  bin_width = 100
  ) +
  xlab(label = "Read lenght (bp)")
```

Plot UMI cluster size distribution.
```{r echo=TRUE, message=FALSE, warning=FALSE}
lu_plot_hist(qc$umi_cluster_size, 1) +
  xlab(label = "UMI cluster size (n reads)")
```

Plot UMI bin size distribution
```{r echo=TRUE, message=FALSE, warning=FALSE}
lu_plot_hist(
  qc$umi_bin_size,
  1
) + 
  xlab(label = "UMI bin size (n reads)")
```

UMI Bin size statistics.
```{r echo=TRUE, message=FALSE, warning=FALSE}
summary(qc$umi_bin_size)
```

Ratio between UMI cluster size and UMI bin size. Media around 4.
```{r echo=TRUE, message=FALSE, warning=FALSE}
lu_plot_hist(
  qc$umi_bin_size/qc$umi_cluster_size,
  1
)
```

### Error profiling

Error rate versus UMI bin size of unfiltered data.
```{r echo=TRUE, message=FALSE, warning=FALSE}
lu_artefact_plot(
  qc,
  breaks = c(seq(1,60,3), Inf)
  ) +
  ylab("Error rate (%)") +
  xlab("UMI Consensus read coverage")
```

Flag contamination and chimeras. Flagged chimeras and contamination below match example data, for de novo processed data correct UMI names have to be inserted.
```{r echo=TRUE, message=FALSE, warning=FALSE}
cont <- tibble(
  umi = c(
    "umi47838bins",
    "umi177241bins",
    "umi13630bins",
    "umi173605bins",
    "umi159967bins",
    "umi185039bins",
    "umi181256bins",
    "umi6705bins",
    "umi42968bins",
    "umi238547bins",
    "umi199820bins",
    "umi151265bins",
    "umi6094bins",
    "umi193827bins",
    "umi181472bins",
    "umi198347bins",
    "umi234765bins",
    "umi180359bins",
    "umi180080bins",
    "umi17145bins",
    "umi189949bins"
),
  flag = "contamination"
)

chi <- tibble(
  umi = c(
    "umi176051bins",
    "umi12507bins",
    "umi183450bins",
    "umi146629bins"
),
  flag = "chimera"
)
```

Inspect chimeras and contamination. Sequences with "Y" have been flagged as chimeras by uchime2_ref.
```{r echo=TRUE, message=FALSE, warning=FALSE}
lu_errorpos_plot(
  ep,
  filter(qc, umi %in% c(cont$umi, chi$umi)),
  species = "",
  flag = "chimera"
)
```

Plot contamination and chimeras for Bacillus
```{r echo=TRUE, message=FALSE, warning=FALSE}
lu_errorpos_plot(
  ep,
  filter(qc, umi %in% c(cont$umi, chi$umi)),
  species = "Bacillus",
  flag= "chimera"
)
```

Filter data to remove chimeras and contamination
```{r echo=TRUE, message=FALSE, warning=FALSE}
qcf <- filter(
  qc,
  !(umi %in% cont$umi),
  !(umi %in% chi$umi)
)
epf <- filter(
  ep,
  !(sub(";.*", "", qname) %in% cont$umi),
  !(sub(";.*", "", qname)  %in% chi$umi)
)
```

Error rate versus UMI bin size of filtered data.
```{r echo=TRUE, message=FALSE, warning=FALSE}
lu_error_plot(
  qcf
  ) +
  ylab("Error rate (%)") +
  xlab("UMI Consensus read coverage")
```

Plot UMI bin size versus error types of filtered data.
```{r echo=TRUE, message=FALSE, warning=FALSE}
lu_errortype_plot(
  epf,
  ref_hp
  ) +
  ylab("Error rate (%)") +
  xlab("UMI Consensus read coverage")
```

Error rate versus UMI bin size intervals of filtered data
```{r echo=TRUE, message=FALSE, warning=FALSE}
lu_error_plot_tbl(qcf) 
```

Filter data < 25x UMI bin size.
```{r echo=TRUE, message=FALSE, warning=FALSE}
qcff <- filter(
  qcf,
  umi_bin_size >= 25
)
epff <- filter(
  epf,
  sub(";.*", "", qname) %in% qcff$umi,
)
```

Error type summary for >=25x data.
```{r echo=TRUE, message=FALSE, warning=FALSE}
lu_errortype_summary(
  profile = epff,
  digits = 4,
  title = NULL
)
```

Plot error frequency pr position of Salmonella operon 7 for >=25x data.
```{r echo=TRUE, message=FALSE, warning=FALSE}
lu_ref_error_frequency(
  profile= epff,
  species = "Salmonella_7",
  lower_cut_off = 0.01,
  label_cut_off = 0.02,
  label_col = "type",
  ylim = c(0,1)
)
```

Error rate divided by operon for >=25x data.
```{r echo=TRUE, message=FALSE, warning=FALSE}
lu_ref_error_plot(
  profile = epff
) +
  ylim(0,10)
```

